The formal verification of the ctm approach to forcing \thanks{Supported by Secyt-UNC projects 33620180100855CB and 33620180100465CB, and Conicet.}

### 2023-01-04 Independence_CH/axioms_metatheory.tex
Some comments inspired by Tim Chow added
Repeated words
Minor formatting edits

\subsection{IFOL and FOL}
In the axioms \isa{refl, subst, allI, spec, exE, exI,eq\_reflection} there is a constraint
for the \emph{type of} the variables \isa{a, b, x} to be in the class \isa{term\_class}.
\begin{isabelle}
IFOL.FalseE: ⋀P. False ⟹ P\isanewline
IFOL.refl:  (⋀a. a = a)\isanewline
IFOL.subst: (⋀a b P. a = b ⟹ P(a) ⟹ P(b))\isanewline
IFOL.allI:  (⋀P. (⋀x. P(x)) ⟹ ∀x. P(x))\isanewline
IFOL.spec:  (⋀P x. ∀x. P(x) ⟹ P(x))\isanewline
IFOL.exE:   (⋀P R. ∃x. P(x) ⟹ (⋀x. P(x) ⟹ R) ⟹ R)\isanewline
IFOL.exI:   (⋀P x. P(x) ⟹ ∃x. P(x))\isanewline
IFOL.conjI: ⋀P Q. P ⟹ Q ⟹ P ∧ Q\isanewline
IFOL.conjunct1: ⋀P Q. P ∧ Q ⟹ P\isanewline
IFOL.conjunct2: ⋀P Q. P ∧ Q ⟹ Q\isanewline
IFOL.disjE: ⋀P Q R. P ∨ Q ⟹ (P ⟹ R) ⟹ (Q ⟹ R) ⟹ R\isanewline
IFOL.disjI1: ⋀P Q. P ⟹ P ∨ Q\isanewline
IFOL.disjI2: ⋀P Q. Q ⟹ P ∨ Q\isanewline
IFOL.eq\_reflection: (⋀x y. x = y ⟹ x ≡ y)\isanewline
IFOL.iff\_reflection: ⋀P Q. P ⟷ Q ⟹ P ≡ Q\isanewline
IFOL.impI: ⋀P Q. (P ⟹ Q) ⟹ P ⟶ Q\isanewline
IFOL.mp: ⋀P Q. P ⟶ Q ⟹ P ⟹ Q\isanewline%\isanewline
FOL.classical: ⋀P. (¬ P ⟹ P) ⟹ P
\end{isabelle}

\subsection{ZF\_Base}
The following symbols are introduced in this theory:
\begin{isabelle}
axiomatization\isanewline
\ \ \ \      mem :: "[i, i] ⇒ o"  (infixl ‹∈› 50)  \isamarkupcmt{membership relation}\isanewline
  and zero :: "i"  (‹0›)  \isamarkupcmt{the empty set}\isanewline
  and Pow :: "i ⇒ i"  \isamarkupcmt{power sets}\isanewline
  and Inf :: "i"  \isamarkupcmt{infinite set}\isanewline
  and Union :: "i ⇒ i"  (‹⋃\_› [90] 90)\isanewline
  and PrimReplace :: "[i, [i, i] ⇒ o] ⇒ i"
\end{isabelle}
\noindent with these axioms
\begin{isabelle}
ZF\_Base.Pow\_iff: ⋀A B. A ∈ Pow(B) ⟷ A ⊆ B\isanewline
ZF\_Base.Union\_iff: ⋀A C. A ∈ ⋃C ⟷ (∃B∈C. A ∈ B)\isanewline
ZF\_Base.extension: ⋀A B. A = B ⟷ A ⊆ B ∧ B ⊆ A\isanewline
ZF\_Base.foundation: ⋀A. A = 0 ∨ (∃x∈A. ∀y∈x. y ∉ A)\isanewline
ZF\_Base.infinity: 0 ∈ Inf ∧ (∀y∈Inf. succ(y) ∈ Inf)\isanewline
ZF\_Base.replacement: ⋀A P b. ∀x∈A. ∀y z. P(x, y) ∧ P(x, z) ⟶ y = z \isanewline
\ \ \ \ \ \ \ \ \ ⟹ b ∈ PrimReplace(A, P) ⟷ (∃x∈A. P(x, b))
\end{isabelle}

>>>>

\subsection{IFOL and FOL}
In the axioms \isa{refl}, \isa{subst}, \isa{allI}, \isa{spec},
\isa{exE}, \isa{exI}, and \isa{eq\_reflection} there is a constraint
for the \emph{type of} the variables \isa{a, b, x} to be in the class \isa{term\_class}.
\begin{isabelle}
IFOL.FalseE: ⋀P. False ⟹ P\isanewline
IFOL.refl:  (⋀a. a = a)\isanewline
IFOL.subst: (⋀a b P. a = b ⟹ P(a) ⟹ P(b))\isanewline
IFOL.allI:  (⋀P. (⋀x. P(x)) ⟹ ∀x. P(x))\isanewline
IFOL.spec:  (⋀P x. ∀x. P(x) ⟹ P(x))\isanewline
IFOL.exE:   (⋀P R. ∃x. P(x) ⟹ (⋀x. P(x) ⟹ R) ⟹ R)\isanewline
IFOL.exI:   (⋀P x. P(x) ⟹ ∃x. P(x))\isanewline
IFOL.conjI: ⋀P Q. P ⟹ Q ⟹ P ∧ Q\isanewline
IFOL.conjunct1: ⋀P Q. P ∧ Q ⟹ P\isanewline
IFOL.conjunct2: ⋀P Q. P ∧ Q ⟹ Q\isanewline
IFOL.disjE: ⋀P Q R. P ∨ Q ⟹ (P ⟹ R) ⟹ (Q ⟹ R) ⟹ R\isanewline
IFOL.disjI1: ⋀P Q. P ⟹ P ∨ Q\isanewline
IFOL.disjI2: ⋀P Q. Q ⟹ P ∨ Q\isanewline
IFOL.eq\_reflection: (⋀x y. x = y ⟹ x ≡ y)\isanewline
IFOL.iff\_reflection: ⋀P Q. P ⟷ Q ⟹ P ≡ Q\isanewline
IFOL.impI: ⋀P Q. (P ⟹ Q) ⟹ P ⟶ Q\isanewline
IFOL.mp: ⋀P Q. P ⟶ Q ⟹ P ⟹ Q\isanewline%\isanewline
FOL.classical: ⋀P. (¬ P ⟹ P) ⟹ P
\end{isabelle}

\subsection{ZF\_Base}
The following symbols are introduced in this theory:
\begin{isabelle}
axiomatization\isanewline
\ \ \ \      mem :: "[i, i] ⇒ o"  (infixl ‹∈› 50)  \isamarkupcmt{membership relation}\isanewline
  and zero :: "i"  (‹0›)  \isamarkupcmt{the empty set}\isanewline
  and Pow :: "i ⇒ i"  \isamarkupcmt{power sets}\isanewline
  and Inf :: "i"  \isamarkupcmt{infinite set}\isanewline
  and Union :: "i ⇒ i"  (‹⋃\_› [90] 90)\isanewline
  and PrimReplace :: "[i, [i, i] ⇒ o] ⇒ i"
\end{isabelle}
\noindent After the definitions of $\notin$, $\subseteq$, $\isa{succ}$,
and relative quantifications are presented, the following axioms are postulated:
\begin{isabelle}
ZF\_Base.Pow\_iff: ⋀A B. A ∈ Pow(B) ⟷ A ⊆ B\isanewline
ZF\_Base.Union\_iff: ⋀A C. A ∈ ⋃C ⟷ (∃B∈C. A ∈ B)\isanewline
ZF\_Base.extension: ⋀A B. A = B ⟷ A ⊆ B ∧ B ⊆ A\isanewline
ZF\_Base.foundation: ⋀A. A = 0 ∨ (∃x∈A. ∀y∈x. y ∉ A)\isanewline
ZF\_Base.infinity: 0 ∈ Inf ∧ (∀y∈Inf. succ(y) ∈ Inf)\isanewline
ZF\_Base.replacement: ⋀A P b. ∀x∈A. ∀y z. P(x, y) ∧ P(x, z) ⟶ y = z \isanewline
\ \ \ \ \ \ \ \ \ ⟹ b ∈ PrimReplace(A, P) ⟷ (∃x∈A. P(x, b))
\end{isabelle}

### 2023-01-04 Independence_CH/lam_replacement.tex
Minor edits
Minor formatting edits

Faced with this situation, we decided that we needed some sort of
\emph{compositionality} in order to obtain new instances from the ones
already proved: Having Replacement for class functions $F$ and $G$
does not entail immediately replacement under $F\circ G$ (unless you
use one further instance of Separation, and the net gain is zero). The
solution was to postulate for the relevant $F$s, instead of
replacement through $x\mapsto F(x)$, a \emph{lambda replacement}
through $x\mapsto \lb x,F(x)\rb$. The name “lambda” corresponds to the
fact that this type of replacement is equivalent to closure under
$(\lambda x\in A.\ F(x)) \defi \{ \lb x,F(x)\rb : x\in A \}$ for every
$A\in M$.

@@

\newcommand{\lamRepl}[2][x]{#1 \mapsto\langle #1,#2\rangle}
\begin{table}[!h]
\centering
\begin{threeparttable}
\begin{tabular}{r<{\stepcounter{LamReplCount}\theLamReplCount.} >{\hspace{1ex}}l @{\hspace{0.8em}} l}
  \toprule
  \multicolumn{1}{r}{No.} & Name & Instance  \\
  \midrule
  \multicolumn{3}{@{}l}{\hspace{0.3em}{\textit{{Replacement Instances}}}}\\
  & \isa{lam{\uscore}replacement{\uscore}fst} & $\lamRepl{\mathit{fst}(x)}$ \\
  & \isa{lam{\uscore}replacement{\uscore}snd} & $\lamRepl{\mathit{snd}(x)}$ \\
  & \isa{lam{\uscore}replacement{\uscore}Union} & $\lamRepl{\bigcup(x)}$ \\
  & \isa{lam{\uscore}replacement{\uscore}Image} & $\lamRepl{\mathit{fst}(x)``\mathit{snd}(x)}$ \\
  & \isa{lam{\uscore}replacement{\uscore}middle{\uscore}del} &
        $\lamRepl{\langle \mathit{fst}(\mathit{fst}(x)),\mathit{snd}(\mathit{snd}(x)) \rangle}$ \\
  & \isa{lam{\uscore}replacement{\uscore}prodRepl} &
        $\lamRepl{\langle \mathit{snd}(\mathit{fst}(x)),\langle \mathit{fst}(\mathit{fst}(x)),\mathit{snd}(\mathit{snd}(x))\rangle \rangle}$\\
  \midrule
  \multicolumn{3}{@{}l}{\hspace{0.3em}{\textit{{Separation Instances}}}}\\
  & \isa{middle{\uscore}separation} & $\mathit{snd}(\mathit{fst}(x))=\mathit{fst}(\mathit{snd}(x))$ \\
  & \isa{separation{\uscore}fst{\uscore}in{\uscore}snd} & $\mathit{fst}(\mathit{snd}(x)) \in \mathit{snd}(\mathit{snd}(x))$\\
  \bottomrule
\end{tabular}
\caption{Replacement and Separation instances of the locale \isa{M{\uscore}replacement}}
\label{tab:m-repl-instances}
\end{threeparttable}
\end{table}

>>>>

Facing this situation, we decided that we needed some sort of
\emph{compositionality} in order to obtain new instances from the ones
already proved: Having Replacement for class functions $F$ and $G$
does not entail immediately replacement under $F\circ G$ (unless you
use one further instance of Separation, and the net gain is zero). The
solution was to postulate for the relevant $F$s, instead of
replacement through $x\mapsto F(x)$, a \emph{lambda replacement}
through $x\mapsto \lb x,F(x)\rb$. The name “lambda” corresponds to the
fact that this type of replacement is equivalent to closure under
$(\lambda x\in A.\ F(x)) \defi \{ \lb x,F(x)\rb : x\in A \}$ for every
$A\in M$.

@@

\newcommand{\lamRepl}[2][x]{#1 \mapsto\langle #1,#2\rangle}
\begin{table}[!h]
\centering
\begin{threeparttable}
\begin{tabular}{r<{\stepcounter{LamReplCount}\theLamReplCount.} >{\hspace{1ex}}l @{\hspace{0.8em}} p{16.3em}}
  \toprule
  \multicolumn{1}{r}{No.} & Name & Instance  \\
  \midrule
  \multicolumn{3}{@{}l}{\hspace{0.3em}{\textit{{Replacement Instances}}}}\\
  & \isa{lam{\uscore}replacement{\uscore}fst} & $\lamRepl{\mathit{fst}(x)}$ \\
  & \isa{lam{\uscore}replacement{\uscore}snd} & $\lamRepl{\mathit{snd}(x)}$ \\
  & \isa{lam{\uscore}replacement{\uscore}Union} & $\lamRepl{\bigcup(x)}$ \\
  & \isa{lam{\uscore}replacement{\uscore}Image} & $\lamRepl{\mathit{fst}(x)``\mathit{snd}(x)}$ \\
  & \isa{lam{\uscore}replacement{\uscore}middle{\uscore}del} &
        $\lamRepl{\langle \mathit{fst}(\mathit{fst}(x)),\mathit{snd}(\mathit{snd}(x)) \rangle}$ \\
  & \isa{lam{\uscore}replacement{\uscore}prodRepl} &
  $x \mapsto \bigl\langle \mathit{snd}(\mathit{fst}(x)),\bigr.$
  \newline
  $\bigl.\phantom{x \mapsto \langle \mathit{snd}}\langle \mathit{fst}(\mathit{fst}(x)),\mathit{snd}(\mathit{snd}(x))\rangle \bigr\rangle$\\
  \midrule
  \multicolumn{3}{@{}l}{\hspace{0.3em}{\textit{{Separation Instances}}}}\\
  & \isa{middle{\uscore}separation} & $\mathit{snd}(\mathit{fst}(x))=\mathit{fst}(\mathit{snd}(x))$ \\
  & \isa{separation{\uscore}fst{\uscore}in{\uscore}snd} & $\mathit{fst}(\mathit{snd}(x)) \in \mathit{snd}(\mathit{snd}(x))$\\
  \bottomrule
\end{tabular}
\caption{Replacement and Separation instances of the locale \isa{M{\uscore}replacement}}
\label{tab:m-repl-instances}
\end{threeparttable}
\end{table}
